---
unit_title: Tidy evalation
needs: tidy-eval
theme: program
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(tibble.print_min = 5, tibble.print_max = 5)
library(tidyverse)
```

## Introduction

At some point during the quarter, you may have noticed that you were copy-and-pasting the same dplyr snippets again and again. You then might have remembered it's a bad idea to have more than three copies of the same code and tried to create a function. Unfortunately if you tried this, you would have failed because dplyr verbs work a little differently to most other R functions. In this reading, you'll learn exactly what makes dplyr verbs different, and a new set of techniques that allow you to wrap them in functions. The underlying idea that makes this possible is __tidy evaluation__, and is used throughout the tidyverse. 

## Quoted arguments

To understand what makes dplyr (and many other tidyverse functions) different, we need some new vocabulary. In R, we can divide function arguments into two classes:

* __Evaluated__ arguments are the default. Code in an evaluated argument 
  executes the same regardless of whether or not its in a function argument.

* Automatically __quoted__ arguments are special; they behave differently 
  depending on whether or not they're inside a function. You can tell if an
  argument is automatically quoted argument by running the code outside
  of the function call: if you get a different result, it's a quoted argument.
  
Let's make this concrete by talking about two important base R functions that you learned about early in the class: `$` and `[[`. When we use `$` the variable name is automatically quoted; if we try and use the name outside of `$` it doesn't work.

```{r, error = TRUE}
df <- data.frame(
  y = 1,
  var = 2
)

df$y

y
```

Why do we say that `$` automatically quotes the variable name? Well, take `[[`. It evaluates its argument so you have to put quotes around it:

```{r}
df[["y"]]
```

The advantage of `$` is concision. The advantage of `[[` is that you can refer to variables in the data frame indirectly:

```{r}
var <- "y"
df[[var]]
```

In base R, there's no consistent way to "unquote", and hence evaluate, an automatically quoted argument. In other words, there's no way to allow `$` to work indirectly:

```{r}
df$var
```

The tidyverse, however, uses a consistent form of unquoting which gives the concision of automatically quoted arguments, while still allowing us to use indirection. Take `pull()`, the dplyr equivalent to `$`:

```{r}
df %>% pull(y)

var <- quo(y)
df %>% pull(!!var)
```

There are two key components: `quo()` which quotes expressions, and `!!` (pronounced bang-bang) that unquotes them.
Here, we're not going to focus on what they actually do, but instead learn how you apply them in practice.

## Wrapping quoting functions

Let's see how to apply your knowledge of quoting vs. evaluating arguments to write a wrapper around some duplicated dplyr code. Take this hypothetical duplicated dplyr code:

```{r, eval = FALSE}
df %>% group_by(x1) %>% summarise(mean = mean(y1))
df %>% group_by(x2) %>% summarise(mean = mean(y2))
df %>% group_by(x3) %>% summarise(mean = mean(y3))
df %>% group_by(x4) %>% summarise(mean = mean(y4))
```

To create a function we need to perform three steps:

1. Identify what is constant and what we might want to vary, and which 
   varying parts are automatically quoted.

1. Create a function template.

1. Quote and unquote the automatically quoted arguments.

Looking at the above code, I'd say there are three primary things that we might want to vary:

* The input data, which I'll call `df`.
* The grouping variable, which I'll call `group_var`.
* The summary variable, which I'll call `summary_var`.

`group_var` and `summary_var` need to be automatically quoted: they won't work when evaluated outside of the dplyr code.

Now we can create the function template using these names for our arguments. 

```{r}
grouped_mean <- function(df, group_var, summary_var) {
}
```

I then copied in the duplicated code and replaced the varying parts with the variable names:

```{r}
grouped_mean <- function(df, group_var, summary_var) {
  df %>% 
    group_by(group_var) %>% 
    summarise(mean = mean(summary_var))
}
```

This function doesn't work (yet), but it's useful to see the error message we get:

```{r, error = TRUE}
grouped_mean(mtcars, cyl, mpg)
```

The error complains that there's no column called `group_var` - that shouldn't be a surprise, because we don't want to use the variable `group_var` directly; we want to use its contents to refer to `cyl`. To fix this problem we need to perform the final step: quoting and unquoting. You can think of quoting as being infectious: if you want your function to vary an automated quoted argument, you also need to quote the corresponding argument. Then to refer to the variable indirectly, you need to unquote it.

```{r}
grouped_mean <- function(df, group_var, summary_var) {
  group_var <- enquo(group_var)
  summary_var <- enquo(summary_var)
  
  df %>% 
    group_by(!!group_var) %>% 
    summarise(mean = mean(!!summary_var))
}
```

If you have eagle eyes, you'll have spotted that I used `enquo()` here but I showed you `quo()` before. That's because they have slightly different uses: `quo()` captures what you, the function writer types, `enquo()` captures what the user has typed:

```{r}
fun1 <- function(x) quo(x)
fun1(a + b)

fun2 <- function(x) enquo(x)
fun2(a + b)
```

As a rule of thumb, use `quo()` when you're experimenting interactively at the console, and `enquo()` when you're creating a function.

## Theory

To finish off, watch this short video to learn the basics of the underlying theory.

<iframe width="560" height="315" src="https://www.youtube.com/embed/nERXS3ssntw" frameborder="0" allow="autoplay; encrypted-media" allowfullscreen></iframe>
