---
unit_title: Vector functions
needs: [manip-basics]
theme: wrangle
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(tibble.print_min = 5, tibble.print_max = 5)
library(tidyverse)
```

## Vector functions

It's often easy to create a __scalar function__, that is a function, that takes length one input and produces a length one output. You can always turn this into a vectored function by figuring out the appropriate purrr `map_` function. It's also easy to accidentally use a vectored function as if it's a scalar function; doing so makes life harder for yourself than it needs to be. This reading illustrates each problem with an example.

## Letter grades

For example, you might write the following function that converts a numeric grade to a letter grade:

```{r}
grade_1 <- function(x) {
  if (x >= 90) {
    "A"
  } else if (x >= 80) {
    "B"
  } else if (x >= 70) {
    "C"
  } else if (x >= 60) {
    "D"
  } else {
    "F"
  }
}

grade_1(92)
grade_1(76)
grade_1(60)
```

But if you attempt to an entire column of a data frame, you have a problem:

```{r}
df <- tibble(
  score = sample(100, 10, replace = TRUE)
)

df %>%
  mutate(grade = grade_1(score))
```

`if` can only work with a single element at a time, so if `grade()` is given a vector it will only use the first element. You can always work around this problem by using one of the `map_` functions from purrr. In this case, `grade()` returns a character vector so we'd use `map_chr()`:

```{r}
df %>%
  mutate(grade = map_chr(score, grade_1))
```

However, there is often an alternative, more elegant approach by relying on an existing vector function. For example, you can always rewrite a set of nested if-else statements to use `case_when()`:

```{r}
grade_2 <- function(x) {
  case_when(
    x >= 90 ~ "A",
    x >= 80 ~ "B",
    x >= 70 ~ "C",
    x >= 60 ~ "D",
    TRUE    ~ "F"
  )
}

grade_2(seq(0, 100, by = 10))

df %>%
  mutate(grade = grade_2(score))
```

And for this particular case, there's an even more targeted function from base R: `cut()`. Its job is to divided a numeric range into named intervals. You give it a vector of breaks, and a vector of labels, and it produces a factor for you:

```{r}
grade_3 <- function(x) {
  cut(x, 
    breaks = c(-Inf, 60, 70, 80, 90, Inf), 
    labels = c("F", "D", "C", "B", "A"),
    right = FALSE
  )
}

grade_3(seq(0, 100, by = 10))
```

In general, there's no easy way to find out that there's an existing function that will make your life much easier. The best technique is to continually expand your knowledge of R by reading widely; a good place to start are the weekly highlights on <http://rweekly.org/>.

## Matching many patterns

So far when you've used stringr, we've always used a single `pattern`. But imagine you have a new challenge: you have a single string and you want see which of a possible set of patterns it matches:

```{r}
private <- tribble(
  ~ name,  ~ pattern,
  "ssn",   "\\d{3}-\\d{2}-\\d{4}",
  "email", "[a-z]+@[a-z]+\\.[a-z]{2,4}",
  "phone", "\\d{3}[- ]?\\d{3}[- ]?\\d{4}"
)

string <- "My social security number is 231-57-7340 and my phone number is 712-458-2189"
```

You might be tempted to use `map_lgl()`:

```{r}
match <- map_lgl(private$pattern, ~ str_detect(string, pattern = .))
private$name[match]
```

But if you carefully read the documentation for `str_detect()` you'll discover that this is unnecessary because `str_detect()` is vectored oven `pattern`. That means you don't need `map_lgl()`!

```{r}
private$name[str_detect(string, private$pattern)]
```

It's sometimes hard to tell from the documentation whether or not an argument is vectored. If reading the docs doesn't help, just try it with a vector; if it works you'll have learned something new and saved yourself a little typing.

